---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---


```{r}
library(survival)
library(tidyverse)
library(survminer)
```
# Load the data

```{r}
data <- read_csv('data/heart_failure_clinical_records_dataset.csv')
```
```{r}
head(data)
```

```{r}
summary(data)
```
```{r}
data$sex <- factor(data$sex, labels= c("female", "male"))
```
```{r}
data$anaemia <- factor(data$anaemia)
```

```{r}
data$diabetes <- factor(data$diabetes)
data$high_blood_pressure <- factor(data$high_blood_pressure)
data$smoking <- factor(data$smoking)
```

```{r}
# Define the breakpoints for the three levels
breakpoints <- c(-Inf, 30, 45, Inf)
# Divide EF into three levels
data$EF_levels <- cut(data$ejection_fraction, breaks = breakpoints, labels = c("EF ≤ 30", "30 < EF ≤ 45", "EF > 45"))

```

# Kaplan-Meyer estimator
```{r}
fit.KM <- survfit(Surv(time,DEATH_EVENT) ~ 1, data = data)
summary(fit.KM)
```


```{r}
fit.KM
```

Median don't exite  because  96 (32%) patients died due to CHD this is low then 50% of patients 
```{r}

ggsurvplot(fit.KM, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
             
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.30, data=data)
```

```{r}
plot(fit.KM, mark.time = TRUE,
     main = "Kaplan-Meier estimator",
     ylab = "Survival probability",
     xlab = "time (seconds)")

```

CDF:
```{r}
plot(fit.KM, fun = "F")
```



## Male vs. Female  
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ sex , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
           legend.labs=c("Female", "Male"), legend.title="Sex",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.30, data=data)
```
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ sex, data = data)
fit.logrank
```
from survival curves of male and female and using the long rank test we can concluded that the sex has no significant impact  

## Smoking
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ smoking , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="smoking",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```

## blood pressure
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ high_blood_pressure , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="high_blood_pressure",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```
###  The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ high_blood_pressure, data = data)
fit.logrank
```
he p-value is 0.04, which is less than 0.05. Therefore, you can conclude that there is evidence of a statistically significant difference in survival between the two groups based on the presence or absence of high blood pressure


## diabetes
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ diabetes , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="diabetes",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```
## anaemia
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ anaemia , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="anaemia",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```
### The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ anaemia, data = data)
fit.logrank
```
## the impacet of age

age is contuins varablie we need firs discrtizei, over 60 beacuse 60 is median 
```{r}
d_age60 <-
  data |>
  mutate(age60 = factor(age <= 60,
                      labels = c("<=60", ">60")))

table(d_age60$age60)
```
```{r}
data$over60 = d_age60$age60

```


```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ over60 , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="over60",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```
### The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ over60, data = data)
fit.logrank
```
## Ejection Fraction
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ EF_levels , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="anaemia",  
           palette=c("dodgerblue2", "orchid2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```

```{r}
plot(sfit, col = c("black", "gray", "red"))
```


## age and Sex
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ over60 + sex , data = data)
sfit
```

```{r}
plot(sfit, col = c("black", "gray", "red", "blue"), lwd = 2)
```

DATA Source
Ahmad, Tanvir; Munir, Assia; Bhatti, Sajjad Haider; Aftab, Muhammad; Ali Raza, Muhammad (2017). DATA_MINIMAL.. PLOS ONE. Dataset. https://doi.org/10.1371/journal.pone.0181001.s001