---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---


```{r}
library(survival)
library(tidyverse)
library(survminer)
```
# Load the data

```{r}
data <- read_csv('data/heart_failure_clinical_records_dataset.csv')
```
```{r}
head(data)
```

```{r}
summary(data)
```
```{r}
data$sex <- factor(data$sex, labels= c("female", "male"))
```
```{r}
data$anaemia <- factor(data$anaemia)
```

```{r}
data$diabetes <- factor(data$diabetes)
data$high_blood_pressure <- factor(data$high_blood_pressure)
data$smoking <- factor(data$smoking)
```

```{r}
# Define the breakpoints for the three levels
breakpoints <- c(-Inf, 30, 45, Inf)
# Divide EF into three levels
data$EF_levels <- cut(data$ejection_fraction, breaks = breakpoints, labels = c("EF ≤ 30", "30 < EF ≤ 45", "EF > 45"))

```

# Kaplan-Meyer estimator
```{r}
fit.KM <- survfit(Surv(time,DEATH_EVENT) ~ 1, data = data)
summary(fit.KM)
```


```{r}
fit.KM
```

Median don't exite  because  96 (32%) patients died due to CHD this is low then 50% of patients 
```{r}

ggsurvplot(fit.KM, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
             
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.30, data=data)
```

```{r}
plot(fit.KM, mark.time = TRUE,
     main = "Kaplan-Meier estimator",
     ylab = "Survival probability",
     xlab = "time (seconds)")

```

CDF:
```{r}
plot(fit.KM, fun = "F")
```



## Male vs. Female  
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ sex , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
           legend.labs=c("Female", "Male"), legend.title="Sex",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.30, data=data)
```
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ sex, data = data)
fit.logrank
```
from survival curves of male and female and using the long rank test we can concluded that the sex has no significant impact  

## Smoking
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ smoking , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="smoking",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```

## blood pressure
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ high_blood_pressure , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="high_blood_pressure",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Failure Survival", 
           risk.table.height=.28, data=data)
```
###  The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ high_blood_pressure, data = data)
fit.logrank
```
he p-value is 0.04, which is less than 0.05. Therefore, you can conclude that there is evidence of a statistically significant difference in survival between the two groups based on the presence or absence of high blood pressure


## diabetes
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ diabetes , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="diabetes",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```
## anaemia
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ anaemia , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="anaemia",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```
### The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ anaemia, data = data)
fit.logrank
```
## the impacet of age

age is contuins varablie we need firs discrtizei, over 60 beacuse 60 is median 
```{r}
d_age60 <-
  data |>
  mutate(age60 = factor(age <= 60,
                      labels = c("<=60", ">60")))

table(d_age60$age60)
```
```{r}
data$over60 = d_age60$age60

```


```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ over60 , data = data)
sfit
```

```{r}
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
            legend.title="over60",  
           palette=c("dodgerblue2", "orchid2"),
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           risk.table.height=.28, data=data)
```
### The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ over60, data = data)
fit.logrank
```
## Ejection Fraction
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ EF_levels , data = data)
sfit
```

```{r}
ggsurvplot(sfit, pval = TRUE , data = data, 
           legend.title="EF levels",
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           )
```
### The logrank test
```{r}
fit.logrank <- survdiff(Surv(time, DEATH_EVENT) ~ EF_levels, data = data)
fit.logrank
```
Since the p-value is less than 0.05 we reject the null hypothesis, this means  there is a statistically significant difference in survival between the three groups. 
## age and anaemia
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ over60 + anaemia , data = data)
sfit
```

```{r}
ggsurv <- ggsurvplot(sfit,  conf.int = TRUE,data=data,
                     ggtheme = theme_bw())

ggsurv$plot +theme_bw() + 
  theme (legend.position = "right")
```
## high_blood_pressure and anaemia
```{r}
sfit <- survfit(Surv(time, DEATH_EVENT) ~ high_blood_pressure + anaemia , data = data)
sfit
```


```{r}
ggsurvplot(sfit, pval = TRUE , data = data, 
           legend.title="",
           title="Kaplan-Meier Curve for Heart Falure Survival", 
           )
```

```{r}
ggsurv <- ggsurvplot(sfit,  conf.int = TRUE,data=data,
                     ggtheme = theme_bw())

ggsurv$plot +theme_bw() + 
  theme (legend.position = "right")+
  facet_grid('anaemia')
```


# Cox Proportional Hazards Model

```{r}
head(data)
```


```{r}
fit_cph <- coxph(Surv(time, DEATH_EVENT) ~ high_blood_pressure + anaemia + EF_levels +over60, data = data)
fit_cph
```
```{r}
test.ph <- cox.zph(fit_cph)
test.ph
```
```{r}
ggcoxzph(test.ph)
```


DATA Source
Ahmad, Tanvir; Munir, Assia; Bhatti, Sajjad Haider; Aftab, Muhammad; Ali Raza, Muhammad (2017). DATA_MINIMAL.. PLOS ONE. Dataset. https://doi.org/10.1371/journal.pone.0181001.s001